---
title: "SA1 Statistical Computing"
author: 
- Bayquen, Christopher Gilbert A.
- Billones, Cristel Kaye
date: "2024-06-26"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, tidy.opts=list(width.cutoff=60), tidy=TRUE) 
```

Use Monte Carlo Simulation to investigate whether the empirical Type I error rate of the t-test is approximately equal to the nominal significance level \( \alpha \), when the sampled population is non-normal. The t-test is robust to mild departures from normality. Discuss the simulation results for the cases where the sampled population is (i) \( X^2 (1)\), (ii) Uniform(0,2), and (iii) Exponential(rate=1). In each case, test \( H_0: \mu = \mu_0 \) vs \(H_1: \mu \neq \mu_0 \), where \(\mu_0\) is the  mean of \(X^2\) (1), Uniform(0,2), and Exponential(1), respectively.

```{r}
# Estimate the Type-I error for different sample sizes
num <- c(10, 25, 50, 100, 200, 500, 1000) 
# Different number of Monte Carlo simulations
m_values <- c(100, 500, 1000, 10000)      

# Initialize an empty list to store the error tables for different m values
err_tables <- list()

for (m in m_values) {
  err_table <- NULL
  
  for (n in num) {
    cv <- qt(0.975, n-1)
    
    # Estimate the Type-I error for chi-square distribution
    er1 <- mean(sapply(1:m, FUN = function(o) {
      x <- rchisq(n, 1)
      m_x <- mean(x)
      se <- sqrt(var(x))
      
      abs((m_x - 1) * sqrt(n) / se) >= cv
    })) 
    
    # Estimate the Type-I error for uniform distribution
    er2 <- mean(sapply(1:m, FUN = function(o) {
      x <- runif(n, 0, 2)
      m_x <- mean(x)
      se <- sqrt(var(x))
      
      abs((m_x - 1) * sqrt(n) / se) >= cv
    }))   
    
    # Estimate the Type-I error for exponential distribution
    er3 <- mean(sapply(1:m, FUN = function(o) {
      x <- rexp(n, 1)
      m_x <- mean(x)
      se <- sqrt(var(x))
      
      abs((m_x - 1) * sqrt(n) / se) >= cv
    }))  
    
    err_table <- cbind(err_table, c(er1, er2, er3))
  }
  
  colnames(err_table) <- num
  rownames(err_table) <- c("chi(1)", "U(0,2)", "exp(1)")
  
  # Store the error table in the list with the number of simulations as the name
  err_tables[[paste0("m_", m)]] <- as.data.frame(err_table)
}

df_100 <- err_tables[["m_100"]]
df_500 <- err_tables[["m_500"]]
df_1000 <- err_tables[["m_1000"]]
df_10000 <- err_tables[["m_10000"]]

```

```{r}
library(ggplot2)
library(reshape2)

# Visualizing the Type I error rates for m = 100 simulations

# Preparing dataframe for visualization
df_100$Distribution <- rownames(df_100)
df100_melt <- melt(df_100, id.vars = "Distribution", variable.name = "SampleSize", value.name = "TypeIErrorRate")


# Plotting
ggplot(df100_melt, aes(x = SampleSize, y = TypeIErrorRate, color = Distribution, group = Distribution)) +
  geom_line() +
  geom_point() +
  labs(title = "Type I Error Rates for Different Distributions (m = 100 Simulations)",
       x = "Sample Size",
       y = "Type I Error Rate") +
  theme_minimal()

```
